### 2020-05-28 serial.tex
done

\newcommand{\gtime}{\sigma} %% generation interval
\newcommand{\gdist}{g} %% generation-interval distribution
\newcommand{\idist}{\ell} %% incubation period distribution

@@

\begin{flushleft}{
	\Large
	\textbf\newline{
		Cohort-based approach to understanding the roles of generation and serial intervals in shaping epidemiological dynamics
	}
}
\newline
\\
Sang Woo Park\textsuperscript{1,*}
Kaiyuan Sun\textsuperscript{2}
Benjamin M.\ Bolker\textsuperscript{3,4,5}
David Champredon\textsuperscript{6}
David J.\,D.\ Earn\textsuperscript{4,5}
Michael Li\textsuperscript{3}
Joshua S.\ Weitz\textsuperscript{7, 8}
Bryan T.\ Grenfell\textsuperscript{1,2,9}
Jonathan Dushoff\textsuperscript{3,4,5}
\\
\bigskip
\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
\\
\textbf{2} Fogarty International Center, National Institutes of Health, Bethesda, MD, USA
\\
\textbf{3} Department of Biology, McMaster University, Hamilton, ON, Canada
\\
\textbf{4} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
\\
\textbf{5} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
\\
\textbf{6} Department of Pathology and Laboratory Medicine, University of Western Ontario, London, Ontario, Canada
\\
\textbf{7} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{8} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{9} Woodrow Wilson School of Public and International Affairs, Princeton University, Princeton, NJ, USA
\\
\bigskip

@@

\bmb{the next two paragraphs are much improved but they might be able to be improved still further by making the logical as explicit as possible (see ? below for points where I think the connections are not absolutely clear)}
One source of confusion arises from an apparent discrepancy between biological intuition and mathematical theory.
When the epidemic is growing exponentially, the spread of infection can be characterized as a ``renewal process'' \bmb{quotes or italic?} based on previous incidence of infection, the associated generation-interval distribution, and the average infectiousness of an infected individual.
This renewal formulation allows us to link the exponential growth rate of an epidemic $r$ with its reproduction number \RR using the generation-interval distribution \citep{wallinga2007generation}.
By definition, the serial-interval distribution describes the renewal process between symptomatic cases based on their symptom onset dates;
therefore \bmb{why? (because both are renewal processes describing the same underlying system, where all components must be growing at the same rate?)}, we expect both generation- and serial-interval distributions to give identical estimates of \RR based on the observed epidemic growth rate $r$.
In contexts where the distributions are expected to be different, current theory has no explanation for how these differing distributions could provide identical estimates of \RR from $r$. 
Some studies have further suggested that using serial intervals can give different estimates of \RR \citep{britton2019estimation, ganyani2020estimating}.

Here, we resolve this discrepancy by showing that the relevant interval for the renewal framework is what is called the ``forward'' interval, and that the forward serial (but not generation) interval is affected by the rate of growth $r$ during the exponential growth phase  \bmb{? how exactly does this resolve the discrepancy?}
We develop a new framework for understanding serial intervals, as well as any other epidemiological delays, and show that the initial forward serial-interval distribution correctly estimates \RR from $r$.
Conversely, using inaccurately defined serial intervals or failing to account for changes in the observed serial-interval distributions over the course of an epidemic can significantly bias estimates of \RR.
We apply our framework to serial intervals of COVID-19 and lay out several principles to consider in using information about serial intervals and other epidemiological time delays to correctly infer the reproduction number during the early stages of an outbreak.

@@

At the individual level, we can define the expected \bmb{I'm not entirely clear what ``expected'' means in this section; it doesn't seem as though this is an expectation in the mathematical sense; what does it mean precisely? (I don't think you necessarily have to give the definition, but knowing the precise definition  will help clarify your usage of ``expected''). Similarly for ``realized'' \ldots} time distribution between a primary and a secondary event for an average infected individual --- we refer to this distribution as the intrinsic distribution.
For example, the intrinsic incubation period distribution describes the expected time distribution from infection to symptom onset of an average infected individual.
Likewise, the intrinsic generation-interval distribution describes the expected time distribution of infectious contacts made by an average infected individual.
However, the intrinsic time distributions are not always equivalent to the corresponding realized time distributions at the population level (i.e., the distribution of time between realized primary and secondary events that occur during an epidemic).
For example, an infectious contact results in infection only if the contacted individual is susceptible (and has already been infected);
this is one mechanism that drives differences between the realized and intrinsic generation-interval distributions.

At the population level, we model realized time delays between a primary and a secondary event from a cohort perspective.
A primary cohort consists of \emph{all} individuals whose primary event occurred at a given time; 
a secondary cohort is defined similarly based on the secondary events.
For example, when we are measuring incubation periods, a primary cohort consists of all individuals who became infected at time $\psymp$.
Similarly, when we are measuring serial intervals, a primary cohort consists of all infectors who became symptomatic at time $\psymp$.
Then, for each primary cohort at time $\psymp$, we can define the expected time distribution between primary and secondary events.
We refer to this distribution as the forward delay distribution and denote it as $f_\psymp(\tau)$.

@@

Multiple \bmb{or ``Several different''} mechanisms drive the changes in forward and backward delay distributions over time.
Typically, within-individual forward delay distributions are not affected by epidemic dynamics.
Some realized distributions, like incubation period distributions, are expected to be equivalent to their intrinsic distributions and remain invariant over the course of an epidemic.
Other realized distributions, like the distribution of time from symptom onset to testing, may change over the course of an epidemic due to changes in public health policies.
Between-individual forward delay distributions, such as generation- or serial-interval distributions, depend on epidemic dynamics;
for example, forward generation intervals contract during the outbreak because infected individuals are less likely to infect others later in  the epidemic due to susceptible depletion \citep{champredon2015intrinsic} \bmb{behavioural change and interventions [i.e. anything that drives time-dependent decreases in transmission] also drive this process.  I know you mention this later in the caveats; do you want to highlight it a little bit more up front since such a large fraction of the decreasing R(t) in COVID-19 could plausibly be driven by processes other than susceptible depletion \emph{per se}?}

\eref{backward} suggests that backward delay distributions change over time even if their corresponding forward delay distribution does not change:
Backward delay distributions depend on changes in the primary cohort size over time due to conditionality of observations (i.e., they are measured from a cohort of individuals whose secondary events occurred at the same time) \bmb{I think this is very important and deserves a few more words to clarify, e.g. ``i.e., conditioning on individuals whose secondary events have already occurred means that we tend not to observe longer inter-event times, so that the distribution is truncated [censored??]'' (or something like that)?}.
When the cohort size is increasing (or decreasing) over time, we are biased to observe shorter (or longer) delays when we look backward.
In particular, when the incidence is growing exponentially, we can calculate the amount of bias exactly.
Assuming that the forward delay distribution remains constant during the exponential growth phase ($f_p(\tau) \approx f_0(\tau)$), we can substitute $\psize(t) = \psize(0) \exp(rt)$ in \eref{backward} to obtain:
\begin{equation}
b_0(\tau) \propto \exp(-r\tau) f_0(\tau),
\label{eq:backexp}
\end{equation}
\djde{I do not find it helpful to write $\propto$.  It would be clearer to write it out exactly with the part that doesn't depend on $\tau$ in $[]$.  Then you can follow with ``where $r$ is the exponential growth rate and the expression in square brackets is a constant since it does not depend on $\tau$.''  I think it is helpful to see what the constant is and not make the reader think it through.  Alternatively, you could write the equality and then have your $\propto$ and then say afterwards that we write $\propto$ for convenience since the expression in $[]$ is constant wrt $\tau$.}
where $r$ is the exponential growth rate.
For faster epidemics (higher $r$), the backward delay distribution will be shorter.
\djde{Is it correct to refer to the distribution as ``shorter''?  I
  believe you mean squashed towards 0.  I'm blanking on the right phrase.}
\bmb{I'm not sure whether ``distribution'' should be singular or plural in general.  Singular reads better to me most of the time, but do whatever you want as long as it's consistent}
Changes in the backward delay distribution are also by definition affected by the changes in the forward delay distribution (\eref{backward}).
Therefore, the backward delay distribution will generally differ from the forward delay distribution even if we are measuring time delays that are intrinsic to the life history of a disease (e.g., the incubation period).
These ideas apply to all epidemiological delay distributions and generalize the work by \cite{champredon2015intrinsic} who compared forward and backward generation-interval distributions to describe the realized generation intervals from the perspective of an infector and an infectee, respectively, as well as the work by \cite{britton2019estimation} who showed that \eref{backexp} holds for the backward generation-interval distributions.

@@

The serial interval is defined as the time between when an infector becomes symptomatic and when an infectee becomes symptomatic \citep{svensson2007note}.
Previous studies have often expressed serial intervals $\tau$ in the form (\fref{diagram}A):
\begin{equation}
\tau = (\gtime + x_2) - x_1
\end{equation}
where $x_1$ and $x_2$ represent incubation periods of an infector and
an infectee, respectively, and $\gtime$ represents the generation
interval.  These studies assumed that $x_1$ and $x_2$ follow the same
distributions and concluded that the serial and generation intervals
have the same mean
\citep{svensson2007note,klinkenberg2011correlation,champredon2018equivalence,
  britton2019estimation}; however, this conclusion is based on the
implicit assumption that incubation times, $x_1$ and $x_2$, as well as
generation interval, $\gtime$, are intrinsic to individuals (and not
dependent on disease \bmb{I think ``epidemic dynamics'' might be
  better than ``disease dynamics'' generally (search and replace?) ---
  emphasizes that it's a population-level effect} dynamics) ---
something that is true of forward but not backward \djde{distribution
  of} incubation times.  \djde{Above is confusing and my suggested
  change doesn't fix it.  Incubation times \underline{are} associated
  with individuals.  It is the distributions that are
  cohort-dependent.}  We refer to this definition as the intrinsic
serial interval (\fref{diagram}A).

\begin{figure}[!th]
\includegraphics[width=\textwidth]{serial_guide.pdf}
\caption{ \djde{I think this graph could be improved using colour.  In
    particular, how about using a red-filled rather than white-filled
    dot for the reference cohort, and using red text for the words
    ``Reference cohort'' everywhere.  I would also put symbols in the
    annotation of the intervals.  For example: ``Intrinsic generation
    interval $\sigma$'' and ``Intrinsic serial interal
    $\sigma+x_2-x_1$'' (though use $\tau_{\rm g}$ etc.
    as discussed by e-mail).}  \textbf{Illustration of intrinsic,
    forward and backward serial intervals.}  (A) The intrinsic serial
  interval for a cohort of individuals infected at time $\psymp$.  In
  this case, $x_1$ follows the intrinsic incubation period
  distribution; $\gtime$ follows the intrinsic generation-interval
  distribution; and $x_2$ follows the intrinsic incubation period
  distribution.  (B) The forward serial interval for a cohort of
  infectors who became symptomatic at time $\psymp$.  In this case,
  $x_1$ follows the backward incubation period distribution; $\gtime$
  follows the forward generation-interval distribution; and $x_2$
  follows the forward incubation period distribution.  (C) The
  backward serial interval for a cohort of infectees who became
  symptomatic at time $\ssymp$.  In this case, $x_1$ follows the
  forward incubation period distribution; $\gtime$ follows the
  backward generation-interval distribution; and $x_2$ follows the
  backward incubation period distribution.  }
\label{fig:diagram}
\end{figure}

To correctly link realized serial intervals to the renewal process
between cases based on symptom onset dates, we must use the forward
serial interval (i.e., use the perspective of a cohort of infectors
that share the same symptom onset time).  Given that an infector
became symptomatic at time $\psymp$, to calculate the forward serial
interval we first go backward in time to when the infector was
infected, and then forward in time to when the infectee was infected,
and then forward to when they became symptomatic.\djde{I altered that
  sentence} In \fref{diagram}B, we see that $x_1$ follows\djde{I think
  ``is drawn from'' would be better than ``follows'' each time in this
  sentence} the backward incubation period distribution of the cohort
of infectors who became symptomatic at time $\psymp$; $\gtime$ follows
the forward generation-interval distribution of the cohort of
infectors who became infected at time $\psymp - x_1$; and $x_2$
follows the forward incubation period distribution of the cohort of
infectees who became infected at time $\psymp - x_1 + \gtime$.
Likewise, we can define the backward serial interval distribution for
a cohort of infectees who became symptomatic at time $\ssymp$
(\fref{diagram}C).  This conceptual framework \djde{delete
  ``clearly''; try to expunge this from your vocabulary; why insult
  somebody for whom it isn't clear?} demonstrates
that the distributions of $x_1$, $\gtime$, and $x_2$ (and therefore
the distributions of realized serial intervals) depend on the
reference cohort and the corresponding cohort time \bmb{have you
  defined what ``cohort time'' means?}.  \djde{``reference time''
  might be better.  How about saying ``reference cohort, which is
  defined by an event type, temporal direction, and a particular
  reference time''.}

To calculate realized serial-interval distributions, we begin by modeling $\total(\psymp,\ssymp)$: the total number of serial intervals that start (when infectors develop symptoms) at time $\psymp$ and end (when infectees develop symptoms) at time $\ssymp$.
The number of serial intervals between time $\psymp$ and $\ssymp$,
given that the infectors became infected at time $\pinf\le\psymp$ and
the infectees became infected at time $\sinf\le\ssymp$, depends on the
amount of infection that occurs between time $\pinf$ and $\sinf$ as
well as the density of forward incubation periods between $\pinf$ and
$\psymp$ (realized incubation periods of infectors) and between
$\sinf$ and $\ssymp$ (realized incubation periods of infectees):
\djde{why not use $t$ instead of $\alpha$?}
\begin{equation}
\underbrace{\Rc (\pinf)}_{\substack{\text{case} \\ \text{reproduction} \\ \text{number}}} 
\times 
\underbrace{i(\pinf)}_{\text{incidence}} 
\times 
\underbrace{h_{\pinf}(\psymp-\pinf, \sinf - \pinf)}_{\substack{\text{joint density of} \\ \text{the forward incubation} \\ \text{period } p-\pinf \text{ and the forward} \\ \text{generation interval } \sinf-\pinf\\ \text{(of infectors)}}}
\times
\underbrace{\idist_{\sinf}(\ssymp - \sinf)}_{\substack{\text{marginal density of} \\ \text{the forward incubation} \\ \text{period } \ssymp-\sinf \\ \text{(of infectees)}}},
\end{equation}
where the case reproduction number $\Rc (\pinf)$ is defined as the average number of secondary cases caused by a primary case infected at time $\pinf$ over the course of their infection \citep{fraser2007estimating}.
We note that \bmb{I always delete this if possible: maybe needed here though?} it is necessary to describe the forward incubation periods and the forward generation intervals using a joint probability distribution because onset of symptoms and transmission potential jointly depend on the life history of a disease;
for example, if an infected individual can only transmit the disease after symptom onset, the forward generation interval will necessarily be longer than the forward incubation period.

\djde{The term ``total number'' of serial intervals is very peculiar,
  because there are infinitely many.  It is more like a cumulative
  distribution.  I'm not sure what the best jargon is.}

The total number of serial intervals can now be obtained by integrating over all possible infection times for the infector and the infectee:
\begin{equation}
\total (\psymp,\ssymp) = \int_{-\infty}^{\psymp} \int_{\pinf}^{\ssymp} \Rc (\pinf) i(\pinf) h_{\pinf}(\psymp-\pinf, \sinf - \pinf) \idist_{\sinf}(\ssymp - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf.
\end{equation}
Then, the forward serial-interval distribution $f_\psymp(\tau)$ is given by the density of intervals of length $\tau$ starting at \psymp, relative to the total number of serial intervals starting at \psymp: 
\begin{equation}
f_\psymp(\tau) = 
\frac{\total(\psymp, \psymp+\tau)}{\int \total(\psymp, \psymp+\tau') \dtau'}.
\label{eq:fserial}
\end{equation}
\djde{Don't leave out the limits of integration.  Ambiguity isn't
  helpful to anyone.  This stuff is hard.  Be as precise as possible.}
Likewise, the backward serial-interval distribution $b_\ssymp(\tau)$ is given by the density of intervals of length $\tau$ ending at \ssymp, relative to the total number of serial intervals ending at \ssymp: 
\begin{equation}
b_\ssymp(\tau) = 
\frac{\total(\ssymp-\tau, \ssymp)}{\int \total(\ssymp-\tau', \ssymp) \dtau'}.
\label{eq:bserial}
\end{equation}
This framework allows us to understand changes in the realized serial intervals for any deterministic epidemic model.

@@

\begin{table}[!th]
\begin{center}
\begin{tabular}{|l|l|r|}
\hline
Parameter & Values & Source\\
\hline
Mean intrinsic incubation period & 5.5 days & \cite{lauer2020incubation} \\
SD intrinsic incubation period & 2.4 days & \cite{lauer2020incubation} \\
Mean intrinsic generation interval & 5 days & \cite{ferretti2020quantifying} \\
SD intrinsic generation interval & 2 days & \cite{ferretti2020quantifying} \\
\hline
\end{tabular}
\end{center}
\caption{
  \textbf{Parameter values used for simulations.}
  \bmb{note that these are chosen to be characteristic of COVID19 (sort of obvious, but most of the paper so far has been general), to illustrate realistic magnitudes of time-varying/perspective effects in the current pandemic?}
The intrinsic generation-interval distribution is parameterized using a log-normal distribution with log mean $\mu_G=1.54$ and log standard deviation $\sigma_G=0.37$.
The intrinsic incubation period distribution is parameterized using a log-normal distribution with log mean $\mu_I=1.62$ and log standard deviation $\sigma_I=0.42$.
The joint probability distribution is modeled using a multivariate log-normal distribution with correlations (on the log scale) $\rho=-0.5, 0, 0.5$.
}
\end{table}

@@

First, we compare the estimates of the basic reproduction number \Ro based on the initial forward serial-interval distributions and the intrinsic serial-interval distributions (see Supplementary Materials for calculation details.).
In particular, we model the joint probability distribution of the intrinsic incubation periods and generation intervals using a bivariate lognormal distribution with correlation $\rho$ based on parameter estimates for COVID-19 (Table 1).
The initial forward serial-interval distributions $f_0(\tau)$ and the intrinsic generation-interval distribution $\gdist(\tau)$ give identical estimates of \Ro (see \eref{Rgen} and \eref{Rforward}) regardless of the correlation $\rho$ between the intrinsic incubation period and the intrinsic generation interval (\fref{rR}A).
However, using the intrinsic serial-interval distributions that do not account for disease dynamics (i.e., assuming that the backward and the forward incubation period distributions are identical) underestimates \Ro (see \eref{Rintrinsic});
as $r$ increases, \Rintrinsic saturates and eventually decreases due to negative serial intervals (\fref{rR}B).
While the forward serial intervals during the exponential growth phase can also be negative, the proportion of negative intervals are appropriately balanced because faster epidemic growth will lead to shorter backward incubation periods (and therefore a lower proportion of negative serial intervals).

>>>>

\newcommand{\gtime}{\tau_{\rm g}} %% generation interval
\newcommand{\gdist}{g} %% generation-interval distribution
\newcommand{\idist}{\ell} %% incubation period distribution

@@

\begin{flushleft}{
	\Large
	\textbf\newline{
		Cohort-based approach to understanding the roles of generation and serial intervals in shaping epidemiological dynamics
	}
}
\newline
\\
Sang Woo Park\textsuperscript{1,*}
Kaiyuan Sun\textsuperscript{2}
David Champredon\textsuperscript{3}
Michael Li\textsuperscript{4}
Benjamin M.\ Bolker\textsuperscript{4,5,6}
David J.\,D.\ Earn\textsuperscript{5,6}
Joshua S.\ Weitz\textsuperscript{7, 8}
Bryan T.\ Grenfell\textsuperscript{1,2,9}
Jonathan Dushoff\textsuperscript{4,5,6}
\\
\bigskip
\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
\\
\textbf{2} Fogarty International Center, National Institutes of Health, Bethesda, MD, USA
\\
\textbf{3} Department of Pathology and Laboratory Medicine, University of Western Ontario, London, Ontario, Canada
\\
\textbf{4} Department of Biology, McMaster University, Hamilton, ON, Canada
\\
\textbf{5} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
\\
\textbf{6} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
\\
\textbf{7} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{8} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{9} Woodrow Wilson School of Public and International Affairs, Princeton University, Princeton, NJ, USA
\\
\bigskip

@@

One source of confusion arises from an apparent discrepancy between biological intuition and mathematical theory.
When the epidemic is growing exponentially, the spread of infection can be characterized as a \emph{renewal process} based on previous incidence of infection, the associated generation-interval distribution, and the average infectiousness of an infected individual.
This renewal formulation allows us to link the exponential growth rate of an epidemic $r$ with its reproduction number \RR using the generation-interval distribution \citep{wallinga2007generation}.
By definition, the serial-interval distribution describes the renewal process between symptomatic cases based on their symptom onset dates;
since both renewal processes, based on generation- and serial-interval distributions, describe the same underlying system, where all components must be growing at the same rate, we expect both generation- and serial-interval distributions to give identical estimates of \RR based on the observed epidemic growth rate $r$.
In contexts where the distributions are expected to be different, current theory has no explanation for how these differing distributions could provide identical estimates of \RR from $r$. 
Some studies have further suggested that using serial intervals can give different estimates of \RR \citep{britton2019estimation, ganyani2020estimating}.

Here, we resolve this discrepancy by showing that the relevant interval for the renewal framework is what is called the ``forward'' interval, and that the forward serial intervals are different from the ``intrinsic'' serial intervals that previous studies have relied on \citep{svensson2007note,klinkenberg2011correlation,te2013estimating,champredon2018equivalence, britton2019estimation}.
We develop a new framework for understanding serial intervals, as well as any other epidemiological delays, and show that the initial forward serial-interval distribution correctly estimates \RR from $r$.
Conversely, using inaccurately defined serial intervals or failing to account for changes in the observed serial-interval distributions over the course of an epidemic can considerably bias estimates of \RR.
We apply our framework to serial intervals of COVID-19 and lay out several principles to consider in using information about serial intervals and other epidemiological time delays to correctly infer the reproduction number during the early stages of an outbreak.

@@

At the individual level, we can define the time distribution between a primary and a secondary event that we expect to observe for an average infected individual --- we refer to this distribution as the intrinsic distribution.
For example, the intrinsic incubation period distribution describes the expected time distribution from infection to symptom onset of an average infected individual.
Likewise, the intrinsic generation-interval distribution describes the expected time distribution of infectious contacts made by an average infected individual.
However, the intrinsic time distributions are not always equivalent to the corresponding realized time distributions at the population level (i.e., the distribution of time between actual primary and secondary events that occur during an epidemic).
For example, an infectious contact results in infection only if the contacted individual is susceptible (and has not already been infected);
this is one mechanism that distinguishes the realized generation intervals (time between actual infection events) from the intrinsic generation intervals (time between infection and infectious contacts).

At the population level, we model realized time delays between a primary and a secondary event that occurs during an epidemic from a cohort perspective.
A primary cohort consists of \emph{all} individuals whose primary event occurred at a given time; 
a secondary cohort is defined similarly based on the secondary events.
For example, when we are measuring incubation periods, a primary cohort consists of all individuals who became infected at time $\psymp$.
Similarly, when we are measuring serial intervals, a primary cohort consists of all infectors who became symptomatic at time $\psymp$.
Then, for each primary cohort at time $\psymp$, we can define the expected time distribution between primary and secondary events.
We refer to this distribution as the forward delay distribution and denote it as $f_\psymp(\tau)$.

@@

Several different mechanisms drive the changes in forward and backward delay distributions over time.
Typically, within-individual forward delay distributions are not affected by epidemic dynamics.
Some realized distributions, like incubation period distributions, are expected to be equivalent to their intrinsic distributions and remain invariant over the course of an epidemic.
Other realized distributions, like the distribution of time from symptom onset to testing, may change over the course of an epidemic due to changes in public health policies.
Between-individual forward delay distributions, such as generation- or serial-interval distributions, depend on epidemic dynamics;
for example, forward generation intervals contract during the outbreak because infected individuals are less likely to infect others later in the epidemic due to factors that drive time-dependent decreases in transmission such as susceptible depletion, behavioral change, and interventions \citep{champredon2015intrinsic}.

\eref{backward} suggests that backward delay distributions change over time even if their corresponding forward delay distribution does not change.
Backward delay distributions depend on changes in the primary cohort size over time due to conditionality of observations:
Conditioning on individuals whose secondary events have occurred at the same time means that the distribution is truncated so that we tend to observe shorter (or longer) inter-event delays when incidence is  the cohort size has been increasing (or decreasing) over time.
In particular, when the incidence is growing exponentially, we can calculate the amount of bias exactly.
Assuming that the forward delay distribution remains constant during the exponential growth phase ($f_p(\tau) \approx f_0(\tau)$), we can substitute $\psize(t) = \psize(0) \exp(rt)$ in \eref{backward} to obtain:
\begin{equation}
b_0(\tau) = \frac{\exp(-r\tau) f_0(\tau)}{\left[\int_{-\infty}^{\infty} \exp(-r\tau') f_0(\tau') \dtau'\right]},
\label{eq:backexp}
\end{equation}
where $r$ is the exponential growth rate and the expression in square brackets is a constant since it does not depend on $\tau$.
For faster epidemics (higher $r$), the backward delay distribution will have a shorter mean.
Changes in the backward delay distribution are also by definition affected by the changes in the forward delay distribution (\eref{backward}).
Therefore, the backward delay distribution will generally differ from the forward delay distribution even if we are measuring time delays that are intrinsic to the life history of a disease (e.g., the incubation period).
These ideas apply to all epidemiological delay distributions and generalize the work by \cite{champredon2015intrinsic} who compared forward and backward generation-interval distributions to describe the realized generation intervals from the perspective of an infector and an infectee, respectively, as well as the work by \cite{britton2019estimation} who showed that \eref{backexp} holds for the backward generation-interval distributions.

@@

The serial interval is defined as the time between when an infector becomes symptomatic and when an infectee becomes symptomatic \citep{svensson2007note}.
Previous studies have often expressed serial intervals $\tau_{\rm s}$ in the form (\fref{diagram}A):
\begin{equation}
\tau_{\rm s} = (\gtime + \tau_{\rm i2}) - \tau_{\rm i1}
\end{equation}
where $\tau_{\rm i1}$ and $\tau_{\rm i2}$ represent incubation periods of an infector and
an infectee, respectively, and $\gtime$ represents the generation
interval.  
These studies assume that $\tau_{\rm i1}$ and $\tau_{\rm i2}$ are drawn from the same distributions and concluded that the serial and generation intervals have the same mean \citep{svensson2007note,klinkenberg2011correlation,champredon2018equivalence, britton2019estimation};
however, this conclusion is based on the implicit assumption that distributions of realized incubation periods, $\tau_{\rm i1}$ and $\tau_{\rm i2}$, as well as generation interval, $\gtime$, are intrinsic to individuals (and not dependent on epidemic dynamics at the population-level) ---
something that is true of forward but not backward distribution of incubation periods.
We refer to this definition as the intrinsic
serial interval (\fref{diagram}A).

\begin{figure}[!th]
\includegraphics[width=\textwidth]{serial_guide.pdf}
\caption{\textbf{Illustration of intrinsic,
    forward and backward serial intervals.}  (A) The intrinsic serial
  interval for a cohort of individuals infected at time $\psymp$.  In
  this case, $\tau_{\rm i1}$ is drawn from the intrinsic incubation period
  distribution; $\gtime$ is drawn from the intrinsic generation-interval
  distribution; and $\tau_{\rm i2}$ is drawn from the intrinsic incubation period
  distribution.  (B) The forward serial interval for a cohort of
  infectors who became symptomatic at time $\psymp$.  In this case,
  $\tau_{\rm i1}$ is drawn from the backward incubation period distribution; $\gtime$
  is drawn from the forward generation-interval distribution; and $\tau_{\rm i2}$
  is drawn from the forward incubation period distribution.  (C) The
  backward serial interval for a cohort of infectees who became
  symptomatic at time $\ssymp$.  In this case, $\tau_{\rm i1}$ is drawn from the
  forward incubation period distribution; $\gtime$ is drawn from the
  backward generation-interval distribution; and $\tau_{\rm i2}$ is drawn from the
  backward incubation period distribution.}
\label{fig:diagram}
\end{figure}

To correctly link the realized serial interval distribution to the renewal process
between cases based on symptom onset dates, we must use the forward
serial interval (i.e., use the perspective of a cohort of infectors
that share the same symptom onset time).  Given that an infector
became symptomatic at time $\psymp$, to calculate the forward serial
interval we first go backward in time to when the infector was
infected, and then forward in time to when the infectee was infected,
and then forward to when the infectee became symptomatic.
In \fref{diagram}B, we see that $\tau_{\rm i1}$ is drawn from
the backward incubation period distribution of the cohort
of infectors who became symptomatic at time $\psymp$; $\gtime$ is drawn from
the forward generation-interval distribution of the cohort of
infectors who became infected at time $\psymp - \tau_{\rm i1}$; and $\tau_{\rm i2}$
is drawn from the forward incubation period distribution of the cohort of
infectees who became infected at time $\psymp - \tau_{\rm i1} + \gtime$.
Likewise, we can define the backward serial interval distribution for
a cohort of infectees who became symptomatic at time $\ssymp$
(\fref{diagram}C).  This conceptual framework demonstrates
that the distributions of $\tau_{\rm i1}$, $\gtime$, and $\tau_{\rm i2}$ (and therefore
the distributions of realized serial intervals) depend on the
reference cohort, which is defined by an event type (primary or secondary),
temporal direction (forward or backward), and a particular reference time.

To calculate realized serial-interval distributions, we begin by modeling $\total(\psymp,\ssymp)$: the total amount of serial intervals that start (when infectors develop symptoms) at time $\psymp$ and end (when infectees develop symptoms) at time $\ssymp$.
The amount of serial intervals between time $\psymp$ and $\ssymp$,
given that the infectors became infected at time $\pinf\le\psymp$ and
the infectees became infected at time $\sinf\le\ssymp$, depends on the
amount of infection that occurs between time $\pinf$ and $\sinf$ as
well as the density of forward incubation periods between $\pinf$ and
$\psymp$ (realized incubation periods of infectors) and between
$\sinf$ and $\ssymp$ (realized incubation periods of infectees):
\begin{equation}
\underbrace{\Rc (\pinf)}_{\substack{\text{case} \\ \text{reproduction} \\ \text{number}}} 
\times 
\underbrace{i(\pinf)}_{\text{incidence}} 
\times 
\underbrace{h_{\pinf}(\psymp-\pinf, \sinf - \pinf)}_{\substack{\text{joint density of} \\ \text{the forward incubation} \\ \text{period } p-\pinf \text{ and the forward} \\ \text{generation interval } \sinf-\pinf\\ \text{(of infectors)}}}
\times
\underbrace{\idist_{\sinf}(\ssymp - \sinf)}_{\substack{\text{marginal density of} \\ \text{the forward incubation} \\ \text{period } \ssymp-\sinf \\ \text{(of infectees)}}},
\end{equation}
where the case reproduction number $\Rc (\pinf)$ is defined as the average number of secondary cases caused by a primary case infected at time $\pinf$ over the course of their infection \citep{fraser2007estimating}.
We describe the forward incubation periods and the forward generation intervals using a joint probability distribution because onset of symptoms and transmission potential jointly depend on the life history of a disease;
for example, if an infected individual can only transmit the disease after symptom onset, the forward generation interval will necessarily be longer than the forward incubation period.

The total amount of serial intervals can now be obtained by integrating over all possible infection times for the infector and the infectee:
\begin{equation}
\total (\psymp,\ssymp) = \int_{-\infty}^{\psymp} \int_{\pinf}^{\ssymp} \Rc (\pinf) i(\pinf) h_{\pinf}(\psymp-\pinf, \sinf - \pinf) \idist_{\sinf}(\ssymp - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf.
\end{equation}
Then, the forward serial-interval distribution $f_\psymp(\tau)$ is given by the density of intervals of length $\tau$ starting at \psymp, relative to the total number of serial intervals starting at \psymp: 
\begin{equation}
f_\psymp(\tau) = 
\frac{\total(\psymp, \psymp+\tau)}{\int_{-\infty}^\infty \total(\psymp, \psymp+\tau') \dtau'}.
\label{eq:fserial}
\end{equation}
Likewise, the backward serial-interval distribution $b_\ssymp(\tau)$ is given by the density of intervals of length $\tau$ ending at \ssymp, relative to the total number of serial intervals ending at \ssymp: 
\begin{equation}
b_\ssymp(\tau) = 
\frac{\total(\ssymp-\tau, \ssymp)}{\int_{-\infty}^\infty \total(\ssymp-\tau', \ssymp) \dtau'}.
\label{eq:bserial}
\end{equation}
This framework allows us to understand changes in the realized serial intervals for any deterministic epidemic model.

@@

\begin{table}[!th]
\begin{center}
\begin{tabular}{|l|l|r|}
\hline
Parameter & Values & Source\\
\hline
Mean intrinsic incubation period & 5.5 days & \cite{lauer2020incubation} \\
SD intrinsic incubation period & 2.4 days & \cite{lauer2020incubation} \\
Mean intrinsic generation interval & 5 days & \cite{ferretti2020quantifying} \\
SD intrinsic generation interval & 2 days & \cite{ferretti2020quantifying} \\
\hline
\end{tabular}
\end{center}
\caption{
  \textbf{Parameter values used for simulations.}
The intrinsic generation-interval distribution is parameterized using a log-normal distribution with log mean $\mu_G=1.54$ and log standard deviation $\sigma_G=0.37$.
The intrinsic incubation period distribution is parameterized using a log-normal distribution with log mean $\mu_I=1.62$ and log standard deviation $\sigma_I=0.42$.
The joint probability distribution is modeled using a multivariate log-normal distribution with correlations (on the log scale) $\rho=-0.5, 0, 0.5$.
The intrinsic generation-interval and incubation period distributions are chosen to match characteristic of COVID-19 to illustrate realistic magnitudes of time-varying/perspective effects in the current pandemic.
}
\end{table}

@@

First, we compare the estimates of the basic reproduction number \Ro based on the initial forward serial-interval distributions and the intrinsic serial-interval distributions (see Supplementary Materials for calculation details.).
In particular, we model the joint probability distribution of the intrinsic incubation periods and generation intervals using a bivariate lognormal distribution with correlation $\rho$ based on parameter estimates for COVID-19 (Table 1).
The initial forward serial-interval distributions $f_0(\tau)$ and the intrinsic generation-interval distribution $\gdist(\tau)$ give identical estimates of \Ro (see \eref{Rgen} and \eref{Rforward}) regardless of the correlation $\rho$ between the intrinsic incubation period and the intrinsic generation interval (\fref{rR}A).
However, using the intrinsic serial-interval distributions that do not account for epidemic dynamics (i.e., assuming that the backward and the forward incubation period distributions are identical) underestimates \Ro (see \eref{Rintrinsic});
as $r$ increases, \Rintrinsic saturates and eventually decreases due to negative serial intervals (\fref{rR}B).
While the forward serial intervals during the exponential growth phase can also be negative, the proportion of negative intervals are appropriately balanced because faster epidemic growth will lead to shorter backward incubation periods (and therefore a lower proportion of negative serial intervals).
